---
title: "Statisitcal Programming - Project 1"
output: rmarkdown::html_document
author: Andrew Amore
---

Due on Friday, October 15th by 5:00 pm (17:00).

Before you start be sure to read through *all* of the rules and instructions in the `README.md`.

<br/>

### Data

```{r}
library(tidyverse)
library(kableExtra)
library(ggplot2)
library(ggpubr)
f1 = readRDS(file="data/f1.rds")
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```


<br/>

### Task 1 - Tidy the data

#### Write up

<!-- Include write up here -->
First, I manually identified the most "granular" nested item from the f1 object: 
races. To correctly format the nested structure I converted races to a tibble 
and leveraged various dplyr functions making use of hoist to cut down on 
unnecessary unnest_wider calls. The top 10 rows were displayed using head.

#### Code

```{r task1}
val = f1$MRData$RaceTable$Races
# convert to tibble
val = tibble::tibble(val)
# dplyr pipeline start
val = val %>% hoist(
  val,
  race_name = "raceName",
  round = "round",
  date = "date",
  results = "Results"
) %>% # elongate results for each race at driver level
  unnest_longer(results) %>%
  # hoist again to extract final table 
  hoist(
    results,
    givenName = c("Driver", "givenName"),
    familyName = c("Driver", "familyName"),
    name = c("Constructor", "name"),
    position = "positionText",
    points = "points"
  ) %>% # concatenate driver name
  mutate(driver = paste(givenName, familyName, sep = " "),
         position = case_when(
          position == 'R' ~ as.character(NA), #"Ret",
          position == 'D' ~ as.character(NA), #"DQ",
          position == 'E' ~ as.character(NA), #"Excluded",
          position == 'W' ~ as.character(NA), #"WD",
          position == 'F' ~ as.character(NA), #"FTQ",
          position == 'N' ~ as.character(NA), #"NC",
          TRUE ~ position
         )) %>%
  select(race_name,
         round,
         date,
         driver,
         name,
         position,
         points) %>%
  rename(constructor = name)
# pull top 10 rows
# head(val, n = 10)

# format and print table
val %>% head(., 10) %>%
  kbl() %>%
  kable_material(c("striped", "hover"))

```

<br/>


### Task 2 - Drivers' Championship

#### Write up

<!-- Include write up here -->
In order to aggregate the correct metric one column, points, was converted
to numeric type. A temporary data structure called total_points was created to
capture total points by driver over the season.

Race position was determined next. The base table from task 1 was pivoted on
race_name to reflect the driver's position in each race. This race table was
joined back to the total_points to create the final task objective.

An additional package, kableExtra, was used to format tables using CSS bootstrap
styling.

#### Code

```{r task2}
# convert column types to match requirements
val$points = as.numeric(val$points)
val$position = as.numeric(val$position)

total_points = val %>%
  group_by(driver) %>%  # get the team results
  summarise(season_points = sum(points))

# join back to the points summary then sort by points...
race_wide = val %>%
  select(race_name, driver, position) %>%
  pivot_wider(names_from = race_name, values_from = position)

# create table df
task2_tbl = race_wide %>% inner_join(total_points, by = "driver") %>%
  arrange(desc(season_points))

# set col names to match wikipedia source
colnames(task2_tbl)[1:ncol(task2_tbl)] <-
  c(
    'Driver',
    'AUT',    'STY',
    'HUN',    'GBR',
    '70A',    'ESP',
    'BEL',    'ITA',
    'TUS',    'RUS',
    'EIF',    'POR',
    'EMI',    'TUR',
    'BHR',    'SKH',
    'ABU',    'Points'
  )

# format and print table
task2_tbl %>%
  kbl() %>%
  kable_material(c("striped", "hover"))
```

<br/>

### Task 3 - Cumulative Constructors

#### Write up

<!-- Include write up here -->
First a temporary data frame was created to assign the final season team 
rankings which will be joined back to final result set. 

Team points were calculated for each race of the season based on driver 
affiliation. After team points, a new column was created to store the cumulative
sum of points throughout the season. This was joined back to the initial temp 
data frame to assign team rankings and final result set sorted based on this
ranking.

#### Code

```{r task3}
# temporary dataframe storing final season team rankings
season_rank = val %>% 
  group_by(constructor) %>%
  summarise(season_point_total = sum(points)) %>%
  mutate(ranking = rank(desc(season_point_total)))

# first compute team points for each race (sum of points for each driver on each
# team).
task3_tbl = val %>% ungroup() %>%
  group_by(race_name, date, constructor) %>%
  summarise(season_points = sum(points)) %>%
  arrange(date) %>% ungroup() %>% group_by(constructor) %>%
  inner_join(season_rank, by = "constructor") %>%
  
  # based on the total team points for each race, compute cumulative sum
  mutate(cum_points = cumsum(season_points)) %>%
  select(-date,-season_points) %>%
  pivot_wider(names_from = race_name, values_from = cum_points) %>%
  arrange(ranking) %>%
  select(-season_point_total,-ranking)

# set col names to match wikipedia source
colnames(task3_tbl)[1:ncol(task3_tbl)] <-
  c(
    'Constructor Name',
    'AUT',    'STY',
    'HUN',    'GBR',
    '70A',    'ESP',
    'BEL',    'ITA',
    'TUS',    'RUS',
    'EIF',    'POR',
    'EMI',    'TUR',
    'BHR',    'SKH',
    'ABU'
  )

# format and print table
task3_tbl %>%
  kbl() %>%
  kable_material(c("striped", "hover"))
```

<br/>

### Task 4 - Visualization

#### Write up

<!-- Include write up here -->

F1 is a truly unique sport combining the marvels of modern engineering
with some of the most gifted reflex-based athletes in the world. Many
people might not be aware of the resource disparities which exist between teams
who compete in the same races for the same championship crown. Just how unequal are 
these teams in terms of the seasonal F1 points system awarded based on final race
position?

This visualization hopes to display how F1 is dominated by the top team(s) by 
displaying the performance of drivers and teams in the 2020 F1 season.

Two charts were created for this purpose:

**Spiral Team Chart (A):**<br/>
Team performance through each race is displayed using a circular bar 
plot. Race abbreviations, denoted along the outside edge, start from 1:00 and 
move clockwise around the circle in order of race occurrence. Points for each
driver are aggregated by team, cumulatively summed over the season, and shown
by color. This chart clearly shows the dominance by team Mercedes over the season
as the color largest block in each race is affiliated with this team. We see this
percentage remain relatively constant from the first few races of the season all
the way to the finale.

**Spiral Driver Chart (B):**<br/>
Point disparities amongst drivers and teams are most magnified at season end. To
show this lopsided distribution a spiral bar chart was created. Drivers are
grouped by constructor and colored according to their affiliation. From this 
chart one can observe just how dominant Lewis Hamilton (and team Mercedes) really
was. Hamilton has almost twice as many points as the next closest opposing driver.
This chart also shows the large number of drivers with 0 points, evident by the
missing axis tick marks denoting point totals.

#### Code

```{r task4, fig.dim = c(10,7)}
############### plot1 - team performance by race (spiral bar) ###############
season = val %>% ungroup() %>%
  group_by(race_name, date, constructor) %>%
  summarise(season_points = sum(points)) %>%
  arrange(date) %>% ungroup() %>% group_by(constructor) %>%
  inner_join(season_rank, by = "constructor") %>%
  mutate(cum_points = cumsum(season_points)) %>%
  select(-season_point_total)

# assign integer denoting order race occurred during season
race_int = val %>% distinct(race_name, date) %>% 
  mutate(race_num = rank(date))%>% select(-date)
team_cir = season %>% inner_join(race_int, by = "race_name") %>% 
  arrange(race_num, cum_points)

# reorder factors for legend ordering
team_cir$constructor <-
  factor(
    team_cir$constructor,
    levels = c(
      "AlphaTauri", "Ferrari",
      "McLaren", "Mercedes",
      "Racing Point", "Red Bull",
      "Renault", "Alfa Romeo",
      "Haas F1 Team", "Williams"
    )
  )

# create data structure for the plot labels
racelabs = c(
  'AUT', 'STY',
  'HUN', 'GBR',
  '70A', 'ESP',
  'BEL', 'ITA',
  'TUS', 'RUS',
  'EIF', 'POR',
  'EMI', 'TUR',
  'BHR', 'SKH',
  'ABU'
)

# plot
fraction_plt = ggplot(team_cir, aes(x = race_num, y = cum_points,
                                    fill = constructor)) +
  geom_col(alpha = 0.5, colour = "black") +
  scale_fill_brewer(palette = "Set3") +
  theme_bw() +
  theme(
    panel.border = element_blank(),
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    plot.title = element_text(hjust = 0.5)
  ) +
  coord_polar() +
  scale_x_continuous(limits = c(1, 18)) +
  guides(fill = guide_legend(title = "Constructor")) +
  xlab("Race Name (In Order of Season)") +
  ylab("Cumulative Points through Each Race")  +
  # set the rotational axis labels to be the race names
  scale_x_discrete(limits = c(seq(
    from = 1, to = 17, by = 1
  )), labels = racelabs)


############# plot2 - individual driver performance (spiral bar) ############# 
# chart inspiration and some source code::
# https://www.r-graph-gallery.com/299-circular-stacked-barplot.html

## stacked bar chart...driver with team
stcked = val %>% group_by(driver, constructor) %>%
  summarise(season_points = sum(points))

data = stcked %>% gather(key = "observation", value = "value", -c(1, 2))

# set ordering for consistency between chart legends
data$constructor <-
  factor(
    data$constructor,
    levels = c(
      "AlphaTauri", "Ferrari",
      "McLaren", "Mercedes",
      "Racing Point", "Red Bull",
      "Renault", "Alfa Romeo",
      "Haas F1 Team", "Williams"
    )
  )

# controls "dart board" arc lengths and label orientation
empty_bar <- 2.5
data <- data %>% arrange(constructor, driver)
data$id <- rep(seq(1, nrow(data) / 1) , each = 1)
# sets label location on chart
label_data <-
  data %>% group_by(id, driver) %>% summarize(tot = sum(value))
number_of_bar <- nrow(label_data)
angle <- 90 - 360 * (label_data$id - 0.5) / number_of_bar
label_data$hjust <- ifelse(angle < -90, 1, 0)
label_data$angle <- ifelse(angle < -90, angle + 180, angle)
# converts data into correct format
base_data <- data %>%
  group_by(constructor) %>%
  summarize(start = min(id), end = max(id) - empty_bar) %>%
  rowwise() %>%
  mutate(title = mean(c(start, end)))

# this sets location for team labels around the interior
grid_data <- base_data
grid_data$end <-
  grid_data$end[c(nrow(grid_data), 1:nrow(grid_data) - 1)] + 1
grid_data$start <- grid_data$start - 1
grid_data <- grid_data[-1,]

driver_plt = ggplot(data) +
  # Add the stacked bar
  geom_bar(
    aes(x = as.factor(id), y = value, fill = constructor),
    stat = "identity",
    alpha = 0.5,
    colour = "black"
  ) +
  scale_fill_brewer(palette = "Set3") +
  
  ## add tick marks (dart board lines) denoting the season point totals
  geom_segment(
    data = grid_data,
    aes(
      x = end,
      y = 0,
      xend = start,
      yend = 0
    ),
    colour = "grey",
    alpha = 1,
    size = 0.1 ,
    inherit.aes = FALSE
  ) +
  geom_segment(
    data = grid_data,
    aes(
      x = end,
      y = 100,
      xend = start,
      yend = 100
    ),
    colour = "grey",
    alpha = 1,
    size = 0.1 ,
    inherit.aes = FALSE
  ) +
  geom_segment(
    data = grid_data,
    aes(
      x = end,
      y = 200,
      xend = start,
      yend = 200
    ),
    colour = "grey",
    alpha = 1,
    size = 0.2 ,
    inherit.aes = FALSE
  ) +
  geom_segment(
    data = grid_data,
    aes(
      x = end,
      y = 300,
      xend = start,
      yend = 300
    ),
    colour = "grey",
    alpha = 1,
    size = 0.3 ,
    inherit.aes = FALSE
  ) +
  geom_segment(
    data = grid_data,
    aes(
      x = end,
      y = 400,
      xend = start,
      yend = 400
    ),
    colour = "grey",
    alpha = 1,
    size = 0.4 ,
    inherit.aes = FALSE
  ) +
  ##
  
  # tick mark labels
  ggplot2::annotate(
    "text",
    x = rep(max(data$id) / 2, 4),
    y = c(100, 200, 300, 400),
    label = c("100", "200", "300", "400") ,
    color = "grey",
    size = 3 ,
    angle = 0,
    fontface = "bold",
    hjust = 1
  ) +
  ggplot2::annotate(
    "text",
    x = rep(max(data$id) / 1.23, 4),
    y = c(100, 200, 300, 400),
    label = c("100", "200", "300", "400") ,
    color = "grey",
    size = 3 ,
    angle = 0,
    fontface = "bold",
    hjust = 1
  ) +
  ggplot2::annotate(
    "text",
    x = rep(max(data$id) / 5.25, 4),
    y = c(100, 200, 300, 400),
    label = c("100", "200", "300", "400") ,
    color = "grey",
    size = 3 ,
    angle = 0,
    fontface = "bold",
    hjust = 1
  ) +
  
  # tick mark range
  ylim(-150, 425) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank()
  ) +
  coord_polar() +
  # Add labels for each bar (driver names)
  geom_text(
    data = label_data,
    aes(
      x = id,
      y = 10,
      label = driver,
      hjust = hjust
    ),
    color = "black",
    fontface = "bold",
    alpha = 0.6,
    size = 3.5,
    angle = label_data$angle,
    inherit.aes = FALSE
  )

# arrange plots
plot = ggarrange(
  fraction_plt,
  driver_plt,
  ncol = 2,
  nrow = 1,
  common.legend = TRUE,
  legend = "top",
  widths = c(1, 1.65),
  heights = c(1, 1.65),
  align = "hv",
  labels = "AUTO"
)
# add main title to figure
annotate_figure(plot,
                top = text_grob(
                  "2020 F1 Driver & Team Performance",
                  color = "black",
                  face = "bold",
                  size = 18
                ))
```

<br/>
